---
title: "FRESHCervid-V6"
author: "DE Spalinger and LA Shipley"
date: "05-05-2026"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Fresh-Cervid model V6
# This is an RStudio version of the FRESH model
# There is no constraint for foraging time in this model
# So, no requirement for bite sizes or NDF concentrations of the foods


#################
#####  NOTES:
################

Before running the model, create a single folder that contains this R program and two additional folders for the data files, and for the outputs, for example:
 "Deer Habitats" and "Deer Habitat Outputs" (without the quotation marks).You can change these folder names to whatever you prefer, but you will also have to change line 43 below to the new name you just created so R can find the data.  In the "Deer Habitats" folder, include the datafiles as specified below.  Each datafile is the data for a single habitat that you wish to examine. However, you can specify as many runs (i.e., changes in constraints on the animals) for each habitat as you wish by adding the new constraints in subsequent rows of the metadata sheet.
 You will also have to modify the folder names for the outputs in lines 202 and 207 as appropriate. 
###############

To run the model, run each shaded block, in sequence, by clicking the green arrow at the right side of the block.  After running the first block (lines 8-10) and the second block (lines 30-36), you no longer need to run these blocks if you want to redo the analyses with modifications of the constraints, etc.  However, redoing the runs will over-write the datafiles in the "Deer Habitat Outputs" folder, so if you wish to keep these, you need to move the files to a different folder before running the program again.


```{r libs}
#Include the necessary library packages:
#install.packages("openxlsx")
#install.packages("lpSolve")
library("lpSolve")
library("openxlsx")
```
#Input Data
#Modify the directories below as you see fit:  "excelpath" is a variable that identifies the directory/folder in which to keep your habitat data.  You may have one or more habitats in this directory, and the program will run each in turn.  It will produce output files that will be placed in a folder you identify/create in line 200 below.

```{r inputs}
# Read in a sample run:
#
excelpath="Deer Habitats/" ## Modify this to reflect the directory with all files to be analyzed
habs = list.files(path=excelpath, pattern="xlsx", all.files=F, full.names=F, recursive=FALSE, ignore.case=T,include.dirs=FALSE)  ## If using convert XLS, ignore.case must be TRUE
# the above writes a character vector of all the files in the above dir
outname3=createWorkbook() # create a new workbook for the summary dataframe of all habitats
addWorksheet(outname3, sheetName="summary")
newsite=c() #create placeholder for site name

```

## Input Specifications
# Sheet 1:
  Sheet 1 Name = "data"
  Nutritional Data required (with units):
    Column 1: Species
    Column 2: Biomass (kg/Ha)
    Column 3: DE (Digestible Energy: kJ/g DM)
    Column 4: DE.SD (standard deviation)
    Column 5: CP (crude Protein: g/100 g DM)
    Column 6: CP.SD (standard deviation)
    Column 7: Max (maximum biomass of plant allowed to be consumed (% of total biomass))
  
  Sheet 2 Name = "metadata"
    Column 1: SiteName (user specified name of habitat)
    Column 2: scenario (user specified name of each scenario to be run)
    Column 3: BodyMass (kg)
    Column 4: MEReq (minimum metabolizable energy requirement of the animal: kJ/day)
    Column 5: CPconst (minimum digestible protein required by the animal: g/100g)
    Column 6: DMIconst (maximum dry matter intake constraint: g DM/day)
    
#. Run the model:
```{r model}
for (m in 1:length(habs)){ # run for each habitat file in the folder
outname=createWorkbook() # create a new workbook for the output dataframe of each habitat run
addWorksheet(outname, sheetName="summary")
excel.file = paste0(c(excelpath,habs[m]),collapse="")
  ## creates a temporary path name for R to find the right file
habitat=loadWorkbook(excel.file)
# read in the plant data for the habitat:
data = data.frame(readWorkbook(excel.file, sheet="data", startRow=1, colNames=TRUE))
# read in the parameters (constraints) for the fresh deer model:
parms=data.frame(readWorkbook(excel.file, sheet="metadata", startRow=1, cols=c(3:6),colNames=TRUE))
#parameters are:  Body Mass, MEReq = metabolizable energy requirement, DMIconst = the dry matter intake constraint,  and CPconst = the crude protein constraint
# with the possibility of running one or more solutions with varying constraints
#ForTime = foraging time constraint (min/day), but this constraint not implimented in this version
site=readWorkbook(excel.file, sheet="metadata", startRow=1, cols=1,colNames=TRUE) # reads in the site name for this run
scenario=readWorkbook(excel.file, sheet="metadata", startRow=1, cols=2,colNames=TRUE) # reads in the Scenarios for this run
olddata=data.frame(data$Species,data$Biomass/3,data$DE,data$CP,data$Max)
colnames(olddata)[1]="Species"
colnames(olddata)[2]="Biomass"
colnames(olddata)[3]="DE"
colnames(olddata)[4]="CP"
colnames(olddata)[5]="Max"
# expand dataset to include variance in nutritional values:
# for each plant in dataset, create 2 new "plants":
spc1=0
spc2=0
de1=0
de2=0
cp1=0
cp2=0
max=0
bio1=0
bio2=0
for(i in 1:nrow(data)){
	spc1[i]=paste(data$Species[i],"-sdev",sep="")
	spc2[i]=paste0(c(data$Species[i],"+sdev"),collapse="")
	de1[i]=data$DE[i]-data$DE.SD[i]
	de2[i]=data$DE[i]+data$DE.SD[i]
	cp1[i]=data$CP[i]-data$CP.SD[i]
	cp2[i]=data$CP[i]+data$CP.SD[i]
	bio1[i]=data$Biomass[i]/3
	bio2[i]=data$Biomass[i]/3
	data$Biomass[i]=data$Biomass[i]/3
	max[i]=data$Max[i]	
}
Species=c(spc1,spc2)
DE=c(de1,de2)
CP=c(cp1,cp2)
Biomass=c(bio1,bio2)
Max=c(max,max)
newdata=data.frame(Species,Biomass,DE,CP,Max)
fdata=rbind(olddata,newdata)
f.obj=fdata$Biomass
totm=0 # this is the max biomass allowed for each forage
dirm=0
conm=matrix(0,nrow=nrow(fdata),ncol=nrow(fdata)) #set the minimum amount allowed for any forage
for(j in 1:nrow(fdata)){
	totm[j]=fdata$Max[j]/100
	dirm[j]="<="
	for(k in 1:nrow(fdata)){
		if(j==k) {
			conm[j,k]=1
		}
		else {conm[j,k]=0}
	}
}
# compute DE and CP constraints:
for(n in 1:nrow(scenario)){	# repeat for different constraint settings in each habitat
  newsite=site$SiteName
de=(parms[n,2]/parms[n,4])/0.82 # converts the ME constraint into a DE constraint, assuming a metabolizable energy coefficient of 0.82
cp=parms[n,3]
bm=parms[n,1]
cde=0
ccp=0
for(l in 1:nrow(fdata)){ #this loop calculates the de and cp for each food relative to the constrained requirements
	cde[l]=fdata$DE[l]*fdata$Biomass[l]-fdata$Biomass[l]*de
	ccp[l]=fdata$CP[l]*fdata$Biomass[l]-fdata$Biomass[l]*cp
}
f.con=c(cde,ccp)
f.con=c(f.con,conm)
f.con=matrix(f.con,nrow=(2+nrow(fdata)), byrow=TRUE)
f.dir=c(">=",">=",dirm)
f.rhs=c(0,0,totm)
out=lp("max",f.obj,f.con,f.dir,f.rhs)$solution # solve the linear equations
#
# compute the summary stats for the run:
#
used=data.frame(out*100)
colnames(used)[1]="Pct_Used"
bioused=data.frame(out*fdata$Biomass)
colnames(bioused)[1]="Biomass_Used"
pctbioused=data.frame((bioused/sum(bioused))*100)
colnames(pctbioused)[1]="Pct_Total_Biomass"
totDE=sum(fdata$DE*bioused)
deave=totDE/sum(bioused)
totCP=sum(fdata$CP*bioused)
cpave=totCP/sum(bioused)
totBio=sum(bioused)
MEintake=deave*parms[n,4]
out2=data.frame(fdata,bioused,used,pctbioused)
out3=data.frame(newsite,sum(fdata$Biomass),totBio,parms[n,1],deave,MEintake,parms[n,2],cpave,parms[n,3],parms[n,4],totBio/(parms[n,4]/1000))
colnames(out3)[1]="Habitat"
colnames(out3)[2]="Total_Biomass"
colnames(out3)[3]="Biomass_Used"
colnames(out3)[4]="Body Weight"
colnames(out3)[5]="Ave DE"
colnames(out3)[6]="Ave MEintake"
colnames(out3)[7]="ME Const"
colnames(out3)[8]="Ave CP"
colnames(out3)[9]="CP Const"
colnames(out3)[10]="DM Intake"
colnames(out3)[11]="Animal_Days_per_Ha"
outname2=paste("output",n,sep="")
addWorksheet(outname,sheetName=outname2)
writeData(outname,out2,sheet=outname2)
if(n==1){
  writeData(outname,out3,sheet="summary",startRow=1,startCol=1)
  if(m==1){
    out4=out3
  }else{
    out4=rbind(out4,out3)
  }
}else
{
  out4=rbind(out4,out3)
  writeData(outname,out3,sheet="summary",startRow=n+1,startCol=1,colNames=FALSE)
}
} #end scenario loop
site2=paste0(c(site,".xlsx"),collapse="")
result=paste0(c("Deer Habitat Outputs/",site2),collapse="")
saveWorkbook(outname,result,overwrite=TRUE)
} #end habitat loop
#
writeData(outname3,out4,sheet="summary",startRow=1,startCol=1)
saveWorkbook(outname3,"Deer Habitat Outputs/RunSummary.xlsx",overwrite=TRUE)

```

